---
title: "Simple LC-MS/MS glycan annotation with GlycoAnnotateR (single file)"
author: "Margot Bligh"
date: "2024-12-11"
output:
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Glycan annotation with GlycoAnnotateR}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This is a *simple* tutorial for the annotation of glycans in a single LC-MS/MS file in R using [XCMS](https://sneumann.github.io/xcms/index.html) and GlycoAnnotateR. For a more detailed tutorial, including correspondence analysis and isotope detection, see this [tutorial](https://margotbligh.github.io/GlycoAnnotateR/lcms_annotation_tutorial.html).

The data used in this tutorial is an enzyme digest of fucoidan, a sulfated fucan from brown algae. The data can be downloaded from the MassIVE server (f.MSV000095410/peak/MS31_20240618_newMpyrifera_digest_SIMddMS2_2.5uL_85%EtOH_11.mzML).


# Load required packages

```{r import packages, message=FALSE, warning=FALSE}
library(GlycoAnnotateR)
library(tidyverse)
library(xcms)
library(ggplot2)
library(ggrepel)
```

# Pre-processing the data
## Import data

Now we import the mzML file as an OnDiskMSnExp object. 


```{r import data}
#read in data 
data <- readMSData(files = 'MS31_20240618_newMpyrifera_digest_SIMddMS2_2.5uL_85%EtOH_11.mzML',
                   mode = 'onDisk')
```

## Peak picking

Peak picking is done with the CentWave algorithm, for details see [Tautenhahn et al., 2008](https://bmcbioinformatics.biomedcentral.com/articles/10.1186/1471-2105-9-504).

*Important to note: each dataset will require parameters appropriate for your data! If no peaks are picked, then you will not be able to annotate any peaks! See this [tutorial](https://bioconductor.org/packages/release/bioc/vignettes/xcms/inst/doc/xcms.html) for pick peaking with CentWave.*

```{r pick peaks}
#create parameter object
cwp <- CentWaveParam(ppm = 10, peakwidth = c(10, 60), integrate = 2,
                     snthresh = 1)
#pick peaks
pks <- findChromPeaks(data, cwp)
```


Peak picking should always be optimised and checked extensively, as this will have a massive impact on the downstream data interpretation. Here, we will check an example chromatogram for a hexasulfated trifucan oligomer  We will quickly use glycoPredict from GlycoAnnotateR to get the correct *m/z* value.

```{r extract chr, warning=FALSE, message=FALSE}
#calculate m/z values 
#first make parameter object
param = glycoPredictParam(dp = c(3,3), #degree of polymerisation
                          polarity = 'neg', #polarity
                          adducts = 'H', # adducts
                          modifications = c('sulfate', #modifications
                                            'deoxy'),
                          double_sulfate = T, #allow double sulfation 
                          nmod_max = 3) #allow 3 modifications per monomer i.e. 2 sulfates, 1 deoxy
#make table
masses_trifucan <- glycoPredict(param)

#inspect table
#there are 82 possible ions 
head(masses_trifucan)

#get mz value for [M-2H]-2 adduct
mz_trifucan = masses_trifucan$mz[masses_trifucan$`IUPAC name` == 'DeoxyHex3 Sulfate6'&
                                   masses_trifucan$ion == '[M-2H]-2']

#extract chromatogram +- 4 ppm
chr_trifucan <- chromatogram(pks,
                             mz = c(mz_trifucan-ppm_to_mz(mz_trifucan, 4), 
                                    mz_trifucan+ppm_to_mz(mz_trifucan, 4)))

```

The picked peak is indicated by the shaded grey area in the extracted ion chromatogram.

```{r plot chr, warning=FALSE, message = FALSE, fig.show='hold', fig.width=6, fig.height=3}
plot(chr_trifucan[[1]])
```

# Annotating peaks

We will now move forward with extracting and annotating our peak table table to identify putative oligosaccharide ions. Here, we will first create the `glycoPredictParam` object, then make a table of potential oligosaccharide compositions. We will filter this for charges of -1 and -2, and then use it to annotate the peak table. The charge filtering may not be appropriate depending on your instrument and dataset. 

## Extract the peak table

We will extract a dataframe with the picked chromatographic peaks

```{r extract peak table}
#extract table and make dataframe
pks_df <- chromPeaks(pks) %>% as.data.frame()

#inspect table
#54 peaks were identified in this example
str(pks_df)
head(pks_df)
```


## Making the parameter object

Create the parameter object for calculation of theoretical compositions.

```{r creating the parameter object}
#create the glycoPredictParam object
gpp <- glycoPredictParam(
  #degree of polymerisation range
  dp = c(1,10), 
  #ionisation polarity
  polarity = "neg", 
  #ionisation type (affects adduct types)
  ion_type = "ESI", 
  #scan range during MS1 acquisition 
  scan_range = c(175,1400),
  #expected modifications
  modifications = c('sulfate', 'deoxy'),
  #double sulfation is possible
  double_sulfate = T,
  #maximum average number of modifications per monomer
  nmod_max = 3,
  #adducts
  adducts = "H")
gpp
```

## Calculating compositions

We can now use parameters defined in the `glycoPredictParam` object to 'predict' (or rather calculate) all possible glycans. You can directly run `glycoAnnotate` to annotate your peak list using the parameters, which will call `glycoPredict` in the background, but here we will demonstrate the output of `glycoPredict` by running it separately, as we want to introduce an addition filtering for charge. You can see that the charge filtered table contains ~1000 ions, while the original table contains >6000.

```{r build calculated df, warning=FALSE}
#create a table of theoretical glycans and their ions
pred_df <- glycoPredict(param = gpp)

#inspect table
str(pred_df)

#charge filter
pred_df_filt <- pred_df %>% filter(charge %in% c('-1', '-2'))

#inspect filtered table
str(pred_df_filt)

```

## Annotate peaks

We will use the charge filtered table to annotate our peaks at 3.5 ppm error.
If you want to 'collapse' annotations in cases where there are multiple annotations per peak (and your peaks are therefore duplicated), use the `collapse = TRUE` option.

```{r annotate peaks}
#annotate peaks
pks_df_annotated <- glycoAnnotate(data = pks_df,
                                  pred_table = pred_df_filt,
                                  error = 3.5)
#keep only annotated peaks in the table
pks_df_onlyannotated <- pks_df_annotated %>% drop_na(dp)

#inspect annotated peaks
#11 sulfated fucan oligosaccharides were annotated
pks_df_onlyannotated
```


# Extract and annotate MS2 spectra associated with annotations
## Extract MS2 spectra

We will use the `glycoMS2Extract` function from GlycoAnnotateR to extract the spectra. This function is basically a wrapper for XCMS functions for MS2 spectra extraction, with a layer to only extract those spectra associated with annotated peaks or features

```{r extract ms2 spectra}
ms2 <- glycoMS2Extract(
  #object with MS1 and MS2 data 
  data_ms2 = data, 
  #processed MS1 data object (must have peaks)
  data_features = pks, 
  #annotated feature table
  #must contain an 'mz' and 'rt' column
  annotations = pks_df_onlyannotated,
  #processing level here is 'peaks' (would be features after correspondence)
  processing_level = 'peaks')

#see how many spectra were extracted in total (here = 961)
length(ms2@backend@spectraData@rownames)
```


## Process MS2 spectra

We will use the functions from `Spectra` to average the MS2 spectra.

```{r process ms2 spectra}
#combine peaks with spectra
ms2_combine <- Spectra::combinePeaks(ms2, tolerance = 0.005, ppm = 3)

#combine spectra by peak within each file across scans
ms2_mean <- Spectra::combineSpectra(ms2_combine, f = ms2$peak_index,
                                    tolerance = 0.005, ppm = 3,  
                                    peaks = 'intersect', minProp = 0.2,
                                    intensityFun = max)

#see how many spectra there are after averaging - 11
length(ms2_mean)
```


## Convert to dataframe

Now we will convert MS2 data into a dataframe format so its easier to work with.


```{r convert spectra to df}
#build empty df
ms2_df <- data.frame(precursorMz = as.numeric(),
                     rt = as.numeric(),
                     mz = as.numeric(),
                     intensity = as.numeric(),
                     peak_id = as.numeric())
#fill in df
for (i in 1:length(ms2_mean)){
  mz = sprintf("%.4f",mz(ms2_mean)[i] %>% unlist()) %>% 
    as.numeric()
  intensity = intensity(ms2_mean)[i] %>% unlist()
  intensity = (intensity / max(intensity)) * 100 #normalise intensity with respect to maximum
  rt = rep(rtime(ms2_mean)[i], length(mz))
  precursorMz = rep(precursorMz(ms2_mean)[i], length(mz))
  peak_id = rep(ms2_mean$peak_id[i], length(mz)) 

  temp <- data.frame(precursorMz = precursorMz,
                     rt = rt, mz = mz, intensity = intensity,
                     peak_id = peak_id)
  ms2_df <- rbind(temp, ms2_df)
}

#inspect dataframe
head(ms2_df)
```



## Annotate MS2 spectra

### Annotate precursors

We will begin by annotating the precursors. 


```{r annotate precursors}
ms2_df_annotP <- glycoAnnotate(data = ms2_df,
                               #this should generally be higher than for MS1
                               error = 5, 
                               #table for annotation
                               pred_table = pred_df_filt, 
                               #name of mz column to annotate) 
                               mz_column = 'precursorMz') 
#make column named 'annotation' with name, ion and dp pasted together
#separated by a colon 
ms2_df_annotP <- ms2_df_annotP %>% 
  mutate(annotations = paste(`IUPAC name`, `ion`, dp, sep = ':'))


```

All precursors should be assigned an annotation (as we only extracted MS2 spectra for annotated precursors).

```{r check precursor annotations}
#check all precursors annotated (statement should be FLASE)
any(is.na(ms2_df_annotP$dp))
```


### Annotate fragments
Now we can annotate the fragment ions in MS2 spectra. The `glycoPredictParam` parameters automatically change depending on the precursor annotation. Only fragments resulting from glycosidic bond breakage, dehydration, or loss of modified groups (e.g. desulfation) can currently be annotated by GlycoAnnotateR. 

```{r annotate fragment ions, message=FALSE, warning=FALSE}
#make a vector of the precursor annotations
#the vector should be as long as the number of peaks / the length
#of the ms2 object after averaging (i.e. 11 in this case)
ms2_df_annotP_unique <- ms2_df_annotP %>% 
  distinct(precursorMz, annotations, rt, peak_id)

#rename annotations column to precursorAnnotations
names(ms2_df_annotP)[names(ms2_df_annotP) == 'annotations'] <- 'precursorAnnotations'

#remove other columns added during annotation
ms2_df_annotP <- ms2_df_annotP %>% 
  select(all_of(c(names(ms2_df), 'precursorAnnotations')))

#annotate the fragment ions based on the precursor annotations
ms2_df_annotF <- glycoMS2Annotate(
  #vector of the precursor annotations
  precursorAnnotations = ms2_df_annotP_unique$annotations,
  #separator in between name, ion and dp of precursor annotations
  precursorAnnotations_sep = ':',
  #ms2spectra dataframe with annotated precursors
  ms2spectra = as.data.frame(ms2_df_annotP),
  #ion type
  ion_type = 'ESI',
  #error
  error = 5, error_units = 'ppm',
  #allow double sulfate and dehydrations and 3 annotations per monomer
  double_sulfate = T, nmod_max = 3, dehydrations = T)

#ions with multiple annotations will automatically be 'collapsed'
#in the 'annotations' column
#ions with one annotation will not be collapsed, so we will now 
#coalesce these columns
#and also annotate sulfate
ms2_df_annotF <- ms2_df_annotF %>% 
  mutate(annotations = case_when(is.na(annotations) & !is.na(ion)~ 
                                   paste(`IUPAC name`, ion, sep = ':'),
                                 is.na(annotations) & is.na(ion) ~  NA,
                                 TRUE ~ annotations),
         annotations = case_when(mz > 96.95 & mz < 97 ~ 'HSO4-',
                                 TRUE ~ annotations))
```


Annotated spectra can be plotted with ggplot.

```{r plot annotated ms2, fig.width=4, fig.height=4, fig.show='hold', warning=FALSE}
#plot spectra with annotated fragments
#you can play around with how you plot this
peakIDs <- ms2_df_annotF$peak_id %>% unique()
for(i in 1:length(peakIDs)){
  pkid <- peakIDs[i]
  df <- ms2_df_annotF %>% 
    #filter for peak id
    filter(peak_id == !!pkid)  %>% 
    #filter for 1% relative abundance
    filter(intensity > 1) %>% 
    #add new line break where there are commas for prettier plotting
    mutate(annotations = gsub('\\,', ',\n', annotations))
  
  
  
  p <-  ggplot(df) +
    #plot lines
    geom_segment(aes(x = mz, xend = mz, y = 0, yend= intensity)) +
    #add annotations
    geom_text_repel(aes(label = annotations, x = mz, y = intensity),
                    size = 2, hjust = 0.5, vjust = 0.5, nudge_y = 1,
                    segment.size = 1, segment.colour = 'lightblue', 
                    segment.linetype = 'dashed', min.segment.length = 0.2) +
    #add mz values 
    geom_text(mapping = aes(x = mz, y = intensity + 0.5, label = round(mz, 4)),
              size = 2) +
    #set labels
    labs(x = expression(italic("m/z")), y = "Normalised intensity (a.u.)",
         title = paste0(df$precursorAnnotations[1], ' at ',
                        round(df$rt[1],1), ' sec, precursor m/z=',
                        round(df$precursorMz[1], 4)))+
    theme_minimal()+
    theme(axis.title = element_text(colour = 'black', size = 8, face = 'bold'),
          axis.text =  element_text(colour = 'black', size = 6),
          plot.title =element_text(colour = 'black', size = 8, face = 'bold', 
                                   hjust = 0.5),
          panel.border = element_rect(fill = NA, colour = 'black'),
          panel.grid = element_blank())
  print(p)
}

```




